\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-19 paper.tex
save

In this note, ... \glnote{todo.}

@@

\glnote{Prescribed vs. implicit. See case of non-diff models in Balaji et al.}

@@

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by sampling mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived as described in Eqn.~\ref{eqn:vo-grad},
which gives
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)].
\end{equation}
\glnote{Smarter approach by exploiting the fact that $\partial d / \partial x$ is known exactly.}

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose collection
of generated samples is optimized to approach the real data distribution $p_r$.

@@

\glnote{Implicit generative models.}
\glnote{ABC.}
\glnote{carl~\citep{cranmer2015approximating}.}
\glnote{Wood's papers.}
\glnote{CMA-ES.}

>>>>

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this likelihood-free setup is often
intractable, as simulators rarely provide a way to evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem with variational optimization. Effectively, the adversarial
variational optimization procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the respective
hierarchical model matches the observed data.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In many fields of science, computer simulators are used to describe complex
processes that tie parameters of an underlying theory to high dimensional
observations. In most cases, these implicit generative
models~\citep{2016arXiv161003483M} are specified as forward stochastic processes
that generate data. Because it is usually computationally intractable,
simulators often do not provide a way to directly evaluate the likelihood
function for a given observation,  which motivates a class of algorithms for
likelihood-free inference.

\glnote{State contributions clearly.}

@@

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by evaluating the exact and known gradient of $U_d$ over
mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
\end{equation}
which we can approximate with mini-batches of generated data
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
\end{equation}
for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?}

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
However, this formulation does not necessarily guarantee that the
proposal distribution $q_\psi$ will place its mass arbitrarily tight
around the parameters of interest $\theta^*$. For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \lambda H(q_\psi)
\end{equation}
where $\lambda \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.

% \glnote{
% While the above gradient estimate $\tilde{\nabla}_\psi U_g$ can readily be plugged
% into the adversarial training procedure, it may also exhibit a sampling
% variance that is larger than necessary, hence making the optimization unstable.
% However, one can exploit the fact that $-d(g(\mathbf{z}; \theta); \phi)$
% is a composition where $-d(\tilde{\mathbf{x}})$ is
% known and differentiable with respect to $\tilde{\mathbf{x}}$.
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
% With large mini-batches, both approaches seem to work fine, though the former seems to be less stable.
% }

@@

\glnote{Implicit generative models.}
\glnote{classifier ABC.}
\glnote{carl~\citep{cranmer2015approximating}.}
\glnote{CMA-ES.}
